---
title: "Loop Habit Tracker – Starter Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1  Introduction

This notebook provides starter code for analysing habit data exported from the
[Loop Habit Tracker](https://loophabits.org/) Android application.  Data are
read and manipulated using the
[`data.table`](https://rdatatable.gitlab.io/data.table/) package.

### 1.1  Loop export format

Loop can export habit history as a CSV file via  
*More → Export*.  The default ("Checkmarks") export produces a **wide** table:

| Name | 2024-01-01 | 2024-01-02 | … |
|------|-----------|-----------|---|
| Exercise | 2 | 0 | … |
| Read | 2 | 2 | … |

Cell values:

| Value | Meaning |
|-------|---------|
| 2 | ✅ Completed |
| 1 | ⏭️ Skipped |
| 0 | ❌ Missed |

---

## 2  Load packages and helper functions

```{r load-packages}
library(data.table)
```

```{r source-helpers}
# Source the utility functions from the R/ folder.
# By default knitr sets the working directory to the notebook's location,
# so the path is relative to notebooks/.
source("../R/read_habits.R")
```

---

## 3  Read the data

Place your exported CSV file in `data/raw/` and update the path below.

```{r read-data}
# ------------------------------------------------------------------
# Wide (Checkmarks) format  – most common Loop export
# ------------------------------------------------------------------
# csv_path <- "../data/raw/Checkmarks.csv"
# habits_dt <- read_habits_wide(csv_path)

# ------------------------------------------------------------------
# Alternatively, create a small synthetic dataset to explore the code
# without needing an export file.
# ------------------------------------------------------------------
set.seed(42)
habits_dt <- data.table(
  habit = rep(c("Exercise", "Read", "Meditate"), each = 90),
  date  = rep(seq.Date(as.Date("2024-01-01"), by = "day", length.out = 90), 3),
  value = sample(c(0L, 1L, 2L), 270, replace = TRUE, prob = c(0.3, 0.1, 0.6))
)

habits_dt <- add_status_label(habits_dt)
```

```{r preview}
head(habits_dt)
```

---

## 4  Basic summary statistics

```{r summary-counts}
# Count observations by habit and status
summary_counts <- habits_dt[, .N, by = .(habit, status)]
setorder(summary_counts, habit, -N)

print(summary_counts)
```

```{r completion-rates}
# Completion rate per habit (excluding skipped days from denominator)
completion_rates <- habits_dt[
  status != "skipped",
  .(
    total      = .N,
    completed  = sum(value == 2L),
    rate_pct   = round(mean(value == 2L) * 100, 1)
  ),
  by = habit
]
setorder(completion_rates, -rate_pct)

print(completion_rates)
```

---

## 5  Monthly aggregation

```{r monthly-agg}
# Add year-month column and aggregate
habits_dt[, year_month := format(date, "%Y-%m")]

monthly <- habits_dt[
  status != "skipped",
  .(
    completed = sum(value == 2L),
    possible  = .N,
    rate_pct  = round(mean(value == 2L) * 100, 1)
  ),
  by = .(habit, year_month)
]
setorder(monthly, habit, year_month)

print(monthly)
```

---

## 6  Streak calculation

A **streak** is the number of consecutive days a habit was completed up to the
most recent observation.

```{r streaks}
#' Calculate the current (most recent) streak for a single habit vector.
#'
#' @param values Integer vector ordered chronologically (oldest first).
#' @return Integer scalar – length of the current streak.
current_streak <- function(values) {
  # Work from the most recent day backwards
  rev_vals <- rev(values)
  streak   <- 0L
  for (v in rev_vals) {
    if (v == 2L) {
      streak <- streak + 1L
    } else {
      break
    }
  }
  streak
}

# Compute streaks per habit
streaks <- habits_dt[
  order(habit, date),
  .(current_streak = current_streak(value)),
  by = habit
]

print(streaks)
```

---

## 7  Pivoting back to wide format

Sometimes it is convenient to view the data in the original wide layout.

```{r wide-pivot}
# Cast back to wide: one column per habit, rows = dates
wide_dt <- dcast(
  habits_dt,
  date ~ habit,
  value.var = "status"
)

head(wide_dt)
```

---

## 8  Saving processed data

```{r save-processed, eval=FALSE}
# Save the long-format table to the processed data folder
fwrite(habits_dt, "../data/processed/habits_long.csv")
```

---

## 9  Next steps

* **Visualisation** – use `ggplot2` or `ggcal` to draw a calendar heatmap.
* **Correlation** – do habits cluster together? Use `dcast` + `cor()`.
* **Time-series modelling** – apply interrupted time-series methods if an
  intervention date is known.
* **Targets pipeline** – promote functions from this notebook into `R/` and
  build a reproducible `{targets}` pipeline.
